---
title: "Single-cell TCR-seq data and epitope mapping"
author:
  name: "Slim Fourati"
  affiliation: "Northwestern University"
  email: "slim.fourati@northwestern.edu"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
---


# 1. Introduction
## Learning goals 
By the end of this demo, you should be able to:  
1. read and inspect single-cell TCR sequencing output  
2. define and summarize clonotypes  
3. quantify clonal expansion and repertoire diversity  
4. visualize V/J gene usage and CDR3 properties  
5. connect TCR sequences to known epitopes using public resources  

## Input data
This practical is built around a **public 10x Genomics BEAM-T dataset** generated from human PBMCs. The dataset contains **gene expression and VDJ-T libraries** from sorted CD8 T cells and includes two known peptide specificities:  
- **Influenza (HLA-A*02:01):** `GILGFVFTL`  
- **CMV (HLA-B*07:02):** `TPRVTGGGAM`  
    
Input dataset was download from the 10x Genomics dataset page (https://www.10xgenomics.com/datasets/5k-human-a0201-b0702-pbmcs-beam-t-2-standard) and place in current working directory called `demo_tcr/`.
This demo uses:  
- `filtered_contig_annotations.csv`

## Folder structure
```text
DGP-485/
└── demo_tcr/
    ├── 20260420_DGP845.BCR_TCR_repertoire.Rmd
    └── filtered_contig_annotations.csv
```

## Load packages
```{r load-packages}
suppressPackageStartupMessages(library(package = "knitr"))
suppressPackageStartupMessages(library(package = "scRepertoire"))
suppressPackageStartupMessages(library(package = "patchwork"))
suppressPackageStartupMessages(library(package = "scales"))
suppressPackageStartupMessages(library(package = "igraph"))
suppressPackageStartupMessages(library(package = "ggraph"))
suppressPackageStartupMessages(library(package = "tidyverse"))
```

# 2. Read the TCR data
```{r read-data}
contigsDF <- read.csv("filtered_contig_annotations.csv")
```

Print the header
```{r inspect-data}
head(contigsDF)
```

# 3. Contig summary
Summarize the structure of the repertoire
```{r basic-summary}
contigsDF %>%
  summarize(n_rows       = n(),
            n_cells      = n_distinct(barcode),
            n_productive = sum(as.logical(productive), na.rm = TRUE),
            chains       = paste(sort(unique(chain)), collapse = ","),
            .groups      = "drop")
```

Count the number of contigs per chain.
```{r chain-counts}
contigsDF %>%
  count(chain, sort = TRUE)
```

Plot chain composition.
```{r plot-chain-counts, fig.height=5, fig.width=4}
contigsDF %>%
  count(chain, sort = TRUE) %>%
  ggplot(aes(x = reorder(chain, n), y = n)) +
  geom_col() +
  labs(title = "Contigs per chain", x = "Chain", y = "Number of contigs") +
  theme_bw()
```

# 4. Keep productive TCR alpha/beta chains
Check whether cells have paired alpha and beta chains.
```{r paired-chains}
pairedStatusDF <- contigsDF %>%
  distinct(barcode, chain) %>%
  count(barcode, chain) %>%
  pivot_wider(names_from = chain, values_from = n, values_fill = 0) %>%
  mutate(paired_ab = `TRA` > 0 & `TRB` > 0,
         tra_only  = `TRA` > 0 & `TRB` == 0,
         trb_only  = `TRB` > 0 & `TRA` == 0)

pairedStatusDF %>%
  summarize(total_cells = n(),
            paired_ab   = sum(paired_ab),
            tra_only    = sum(tra_only),
            trb_only    = sum(trb_only),
            .groups     = "drop")
```

# 5. Build a clonotype object with scRepertoire
`scRepertoire` is widely used in teaching because it provides simple functions for clonal frequency, diversity, and repertoire visualization.
```{r combine-tcr}
combinedLS <- combineTCR(
  input.data = list(contigsDF),
  samples = "BEAMT",
  ID = "demo_tcr")
```

Inspect the object.
```{r inspect-combined}
str(combinedLS, max.level = 1)
head(combinedLS[[1]])
```

# 6. Clonal expansion
A central question in repertoire analysis is whether a small number of clonotypes dominate the sample.
```{r clonal-homeostasis-plot}
clonalHomeostasis(combinedLS, cloneCall = "aa")
```

```{r clonal-proportion-plot}
clonalProportion(combinedLS, cloneCall = "aa")
```
Questions:
- Is the repertoire evenly distributed?  
- Do you see evidence for large expanded clones?  
- What biological processes can generate expansion of a clone?  

# 7. Diversity metrics
Diversity is not the same thing as the number of cells. It reflects both the **number of distinct clonotypes** and **how evenly they are represented**.
```{r diversity}
clonalDiverityPlot <- clonalDiversity(combinedLS, cloneCall = "aa")
clonalDiverityPlot@data
```
Which metric would decrease most if one clone became extremely dominant?

# 8. Identify the most expanded clonotypes
Compute the most frequent paired clonotypes directly from the contig table.
```{r top-clones}
topClonesDF <- contigsDF %>%
  filter(!is.na(raw_clonotype_id), raw_clonotype_id != "") %>%
  distinct(barcode, raw_clonotype_id) %>%
  count(raw_clonotype_id, sort = TRUE) %>%
  slice_head(n = 10)

topClonesDF
```

Visualize the top clonotypes.

```{r top-clones-plot}
ggplot(data    = topClonesDF,
       mapping = aes(x = reorder(raw_clonotype_id, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 10 clonotypes",
       x     = "Clonotype ID",
       y     = "Number of cells") +
  theme_bw()
```

# 9. Explore CDR3 sequences
The CDR3 region is the most variable part of the receptor and is usually the main sequence used for specificity-oriented analyses.
```{r cdr3-table}
cdr3TableDF <- contigsDF %>%
  select(barcode, chain, v_gene, j_gene, cdr3, raw_clonotype_id) %>%
  distinct()

head(cdr3TableDF)
```

Plot CDR3 length distributions.
```{r cdr3-length}
cdr3LengthsDF <- cdr3TableDF %>%
  filter(!is.na(cdr3), cdr3 != "") %>%
  mutate(cdr3_length = nchar(cdr3))

ggplot(data = cdr3LengthsDF, mapping = aes(x = cdr3_length)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~chain, scales = "free_y") +
  labs(title = "CDR3 amino-acid length distribution", 
       x     = "CDR3 length", 
      y      = "Count") +
  theme_bw()
```
Do TRA and TRB CDR3 lengths appear similar? Why might they differ?

# 10. V and J gene usage
V and J usage can reflect both recombination biases and antigen-driven selection.
```{r v-gene-usage}
vUsageDF <- cdr3TableDF %>%
  count(chain, v_gene, sort = TRUE) %>%
  group_by(chain) %>%
  slice_max(n, n = 12) %>%
  ungroup()

vUsageDF
```

```{r plot-v-usage}
ggplot(data = vUsageDF, mapping = aes(x = reorder(v_gene, n), y = n)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~chain, scales = "free_y") +
  labs(title = "Top V genes", x = "V gene", y = "Count") +
  theme_bw()
```

Now examine V/J combinations.
```{r vj-heatmap}
vjUsageDF <- cdr3TableDF %>%
  filter(chain == "TRB", !is.na(v_gene), !is.na(j_gene)) %>%
  count(v_gene, j_gene)

ggplot(data = vjUsageDF, mapping = aes(x = j_gene, y = v_gene, fill = n)) +
  geom_tile() +
  scale_fill_continuous(labels = comma) +
  labs(title = "TRB V/J usage", x = "J gene", y = "V gene", fill = "Count") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# 11. From clonotype to epitope: preparing candidate sequences
For epitope mapping, a common starting point is to inspect the most expanded TCRs, especially the **TRB CDR3 amino-acid sequences**.
```{r top-trb-cdr3}
topTrbCdr3DF <- contigsDF %>%
  filter(chain == "TRB", !is.na(cdr3), cdr3 != "") %>%
  distinct(barcode, raw_clonotype_id, cdr3, v_gene, j_gene) %>%
  count(raw_clonotype_id, cdr3, v_gene, j_gene, sort = TRUE) %>%
  slice_head(n = 20)

topTrbCdr3DF
```

Export a short list for manual database interrogation.
```{r export-top-cdr3}
write_csv(topTrbCdr3DF, file = "top_TRB_CDR3_for_epitope_search.csv")
```

# 12. Manual epitope mapping using public databases
Use the table above and query the following public resources:
**IEDB** (https://tools.iedb.org/tcrmatch/result/)  

For each of the top 5 TRB CDR3 sequences, record:
1. whether an exact or near match exists;
2. the reported antigen / epitope;
3. the HLA restriction;
4. the pathogen or disease context.
```{r read-iedb-search-output}
iedbRes <- read_tsv(file = "tcrmatch_result_8381De.tsv")

iedbRes %>% filter(grepl(pattern = "GILGFVFTL", epitope)) %>%
  arrange(desc(score))
filter(topTrbCdr3DF, grepl(pattern = "ASSSANYGYT", cdr3))

iedbRes %>% filter(grepl(pattern = "TPRVTGGGAM", epitope)) %>%
  arrange(desc(score))
filter(topTrbCdr3DF, grepl(pattern = "ASSGLAGLLETQY", cdr3))
```
Do any of your top clonotypes map to influenza or CMV-associated receptors? If not, what are the possible explanations?

# 13. Optional extension: sequence similarity network
A simple way to introduce the idea of receptor similarity is to connect TRB CDR3s that differ by only a small edit distance.
> This is **not** a full specificity model, but it provides intuition for motif clustering methods such as GLIPH2 or TCRdist.
```{r similarity-network-prep}
smallSetDF <- topTrbCdr3DF %>%
  slice_head(n = 20) %>%
  mutate(node_id = row_number())

edgesDF <- expand.grid(i = smallSetDF$node_id, j = smallSetDF$node_id) %>%
  as_tibble() %>%
  filter(i < j) %>%
  left_join(smallSetDF %>% select(i = node_id, cdr3_i = cdr3), by = "i") %>%
  left_join(smallSetDF %>% select(j = node_id, cdr3_j = cdr3), by = "j") %>%
  mutate(dist = mapply(adist, cdr3_i, cdr3_j)) %>%
  filter(dist <= 2)

edgesDF
```

```{r similarity-network-plot}
if (nrow(edgesDF) > 0) {
  g <- graph_from_data_frame(
    d = edgesDF %>% select(i, j, dist),
    directed = FALSE,
    vertices = smallSetDF %>% select(node_id, cdr3, n)
  )

ggraph(g, layout = "fr") +
    geom_edge_link(aes(alpha = 1 / (dist + 1))) +
    geom_node_point(aes(size = n)) +
    geom_node_text(aes(label = cdr3), repel = TRUE, size = 3) +
    labs(title = "TRB CDR3 similarity network (edit distance <= 2)") +
    theme_bw()
} else {
  cat("No edges found at edit distance <= 2 in this subset. Try a larger threshold.\n")
}
```

# Take-home message
TCR repertoire analysis is one of the clearest examples of how sequencing, statistics, and immunology intersect. In a single workflow, students can move from:
- raw VDJ output,
- to clonotype structure,
- to diversity and expansion,
- to hypotheses about antigen specificity.
